Modular Composing Typed Language Fragments

### 2014-12-02 att-pldi15/att-pldi15.tex

In contrast, in @$\lambda$, fragment providers need not provide the semantics with mechanized specfications or proofs to benefit from rigorous modular reasoning principles. Instead, under a fixed tycon context, the calculus can be reasoned about like a very small type-directed compiler \cite{tarditi+:til-OLD,conf/pldi/Chlipala07}. Errors in reasoning can only lead to failure at typechecking time, via our chief contribution: a novel form of \emph{translation validation} \cite{Pnueli-Siegel-Singerman98}. Incorrect opcon definitions relative to a specification (e.g. \cite{sanitation-psp14} for regular strings) can at worst weaken expected invariants at that tycon, like incorrectly implemented modules in ML. Thus, modular tycons can reasonably be tested ``in the field'' without concern about the reliability of the semantics as a whole. 
To eliminate even these localized failure modes, we plan to introduce \emph{optional} specification and proof mechanization into the SL (by basing it on a dependently typed language like Coq, rather than ML). Because types are values in the SL, the SL can be seen as building on the concept of \emph{type-level computation}. %We also hope to mechanize the metatheory of @$\lambda$ itself in future work.%Every combination of fragments is a new dialect, and must be reasoned about monolithically. %Efforts to reason modularly areeed, several problems can come up when fragments are combined, so . 
% Let us briefly review difficulties that arise.

@@

Type abstraction, encouragingly, also underlies modular reasoning in ML-like languages \cite{pfpl,harper1997programming} and languages with other forms of ADTs \cite{liskov1974programming} like Scala \cite{conf/oopsla/AminRO14}. Indeed, proofs of tycon invariants can rely on existing parametricity theorems \cite{WadlerThms}. 
Our work is thus reminiscent of  work on elaborating an ML-style module system into System $\mathbf{F}_\omega$ \cite{conf/tldi/RossbergRD10}.  Unlike in module systems, type translations (analagous to the choice of representation for an abstract type) are statically \emph{computed} based on a type index, rather than statically \emph{declared}. Moreover, there can be arbitrarily many operators because they arise by providing a term index to an opcon, and their  semantics can be complex because a static function computes the types and translations that arise. In contrast, modules and ADTs can only specify a fixed number of operations, and each must have function type. Note also that these are not competing mechanisms: we did not specify quantification over external types here for simplicity, but we conjecture that this is entirely complementary and thus @$\lambda$ could serve as the core of a language with an ML-style module system and parametric polymorphism. Another related direction is to explore \emph{tycon functors}, which would abstract over tycons with the same signature to support modularly tunable cost semantics. % This permits us tocode targeting a separate IL, and permit operations whose types cannot simply be written as functions. %Both examples we used here would be quite difficult to fully embed using only modules.

\section{Conclusion}
\noindent We have specified a remarkably simple but surprisingly powerful typed translation semantics, @$\lambda$, where new type constructors can be introduced ``from within''. The corresponding operator constructors determine types and translations ``actively'', i.e. using static functions. A simple form of translation validation based on existing notions of type abstraction  ensures that opcons associated with one tycon maintain translation independence from all others, guaranteeing that a wide class of important properties can be reasoned about modularly. Implementation is ongoing.

>>>>

In contrast, in @$\lambda$, fragment providers need not provide the semantics with mechanized specfications or proofs to benefit from rigorous modular reasoning principles. Instead, under a fixed tycon context, the calculus can be reasoned about like a very small type-directed compiler \cite{tarditi+:til-OLD,conf/pldi/Chlipala07,Dave:2003:CVB:966221.966235}. Errors in reasoning can only lead to failure at typechecking time, via our chief contribution: a novel form of \emph{translation validation} \cite{Pnueli-Siegel-Singerman98}. Incorrect opcon definitions relative to a specification (e.g. \cite{sanitation-psp14} for regular strings) can at worst weaken expected invariants at that tycon, like incorrectly implemented modules in ML. Thus, modular tycons can reasonably be tested ``in the field'' without concern about the reliability of the semantics as a whole. 
To eliminate even these localized failure modes, we plan to introduce \emph{optional} specification and proof mechanization into the SL (by basing it on a dependently typed language like Coq, rather than ML). Because types are values in the SL, the SL can be seen as building on the concept of \emph{type-level computation}\todo{find appropriate citation}. %We also hope to mechanize the metatheory of @$\lambda$ itself in future work.%Every combination of fragments is a new dialect, and must be reasoned about monolithically. %Efforts to reason modularly areeed, several problems can come up when fragments are combined, so . 
% Let us briefly review difficulties that arise.

@@

Type abstraction, encouragingly, also underlies modular reasoning in ML-like languages \cite{pfpl,harper1997programming} and languages with other forms of ADTs \cite{liskov1974programming} like Scala \cite{conf/oopsla/AminRO14}. Indeed, proofs of tycon invariants can rely on existing parametricity theorems \cite{WadlerThms}. 
Our work is thus reminiscent of  work on elaborating an ML-style module system into System $\mathbf{F}_\omega$ \cite{conf/tldi/RossbergRD10}.  Unlike in module systems, type translations (analagous to the choice of representation for an abstract type) are statically \emph{computed} based on a type index, rather than statically \emph{declared}. Moreover, there can be arbitrarily many operators because they arise by providing a term index to an opcon, and their  semantics can be complex because a static function computes the types and translations that arise. In contrast, modules and ADTs can only specify a fixed number of operations, and each must have function type. Note also that these are not competing mechanisms: we did not specify quantification over external types here for simplicity, but we conjecture, based on the finding that polymorphism is conservative over simple types, \cite{breazu1990polymorphism}, that this is  complementary and thus @$\lambda$ could serve as the core of a language with an ML-style module system and polymorphism. Another related direction is to explore \emph{tycon functors}, which would abstract over tycons with the same signature to support modularly tunable cost semantics. % This permits us tocode targeting a separate IL, and permit operations whose types cannot simply be written as functions. %Both examples we used here would be quite difficult to fully embed using only modules.

\section{Conclusion}
\noindent We have specified a simple but surprisingly powerful typed translation semantics, @$\lambda$, where new type constructors can be introduced ``from within''. The corresponding operator constructors determine types and translations ``actively'', i.e. using static functions. A simple form of translation validation based on existing notions of type abstraction  ensures that opcons associated with one tycon maintain translation independence from all others, guaranteeing that a wide class of important properties can be reasoned about modularly. Implementation in several settings is ongoing.
